---
title: "Reproducible results and discussion subsections for Kaytetye medial vowels paper"
author: "Nay San"
date: "Generated `r format(Sys.time(), '%F %R UTC%z')`"
output:
  html_notebook:
    code_folding: hide
    number_sections: no
    toc: yes
  html_document:
    df_print: kable
    number_sections: no
    toc: yes
  word_document:
    toc: yes
params:
    random_seed: 3
---

```{r Setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(stringsAsFactors = FALSE)    
    
library(mclust)
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(here)
library(DT)    
    
source(here("src/visualization/mclust-plot_helper.R"))
source(here("src/data/mclust_helpers.R"))    
    
vowels_med_analysis <- read_csv(here("data/processed/vowels_med_analysis.csv"), col_types = "ciicccccddcdddddddddd")
micl_results <- read_csv(here("data/processed/micl_results.csv"), col_types = "iciicd")    

# Some helpers
placeholder <- '<span style="color:red">X</span>'

# Set up ggplot theme
bw_custom <- theme_set(
    theme_bw(base_size = 14) +
    theme(
        legend.position="none",
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()
    )
)

# Some helpers
placeholder <- '<span style="color:red">X</span>'
```

# Comparison of all fitted models

Figure `r placeholder` below shows the maxium ICL value for each cluster number over 10,000 iterations; the text labels within the figure show the absolute maximum.
For both Forest- and Praat-tracker formant values, 4 is the optimal number of clusters, marginally better than 3, but considerably better than both 2 and 5 clusters.

```{r Show max. ICL graphs, message=FALSE, fig.width=6, fig.height=2.5, fig.align='center'}
micl_best <-
    micl_results %>%
    group_by(tracker, G) %>%
    filter(ICL == max(ICL)) %>%
    arrange(tracker, G) %>% 
    ungroup %>% 
    mutate(Tracker = str_to_title(tracker)) %>%
    select(Tracker, G, maxICL = ICL, modelName, seed, iteration)

micl_results %>%
    group_by(iteration, tracker, G) %>%
    summarise(maxICL = max(ICL)) %>% 
    ungroup() %>%
    filter(tracker == "praat") %>% 
    # mutate(Tracker = str_to_title(tracker)) %>%
    ggplot(aes(x = as.factor(G), y = maxICL)) +
    ggrastr::geom_point_rast() +
    # facet_wrap(~ Tracker, ncol = 2) +
    ggrepel::geom_label_repel(
        data = filter(micl_best, Tracker == "Praat"),
        aes(label = round(maxICL, digits = 0)),
        arrow = arrow(length = unit(0.03, "npc"), type = "closed", ends = "last"),
        vjust = -2,
        hjust = -0.25
    ) +
    xlab("G (number of clusters)") +
    ylab("Max. ICL") 
    
```
**Figure `r placeholder`** Maximum integrated complete-data likelihood (ICL) values from 10,000 iterations of fitting Gaussian mixture models (GMMs) onto z-normalized mid-point formant first and second formant values, measured using two formant trackers (Forest, Praat). Arrowed text labels annotated the maximum ICL value for G, the number of clusters fitted by the GMM. 

# Comparison between optimal models and human annotators

```{r Setup to show optimal models, message=FALSE, include=FALSE}
group_colors <- c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")
group_shapes <- c(0,1,2,5,6)

mclust_best_results <- with(
    data = micl_best,
    pmap(
        list(str_to_lower(Tracker), G, modelName, seed),
        function(tracker, G, modelName, seed) {
                df <- sample_vowel_midpoints(tracker = tracker, seed = seed)
                names(df)[1:2] <- c("f1_z", "f2_z")
                
                set.seed(seed)
                mclust_fit <- Mclust(
                    data = df[, 1:2],
                    G = G,
                    modelName = modelName
                )
                
                classifications <-
                    cbind(
                        G,
                        tracker, 
                        seed,
                        df,
                        clust = mclust_fit$classification,
                        uncertainty = mclust_fit$uncertainty,
                        stringsAsFactors = FALSE
                    ) %>%
                    left_join(select(vowels_med_analysis, source_file:base_vowel, annotator))
                
                centres <-
                    mclust_fit$parameters$mean %>%
                    cbind(row.names(.), .) %>%
                    as_tibble() %>%
                    set_names(c("formant", 1:mclust_fit$G)) %>%
                    gather(key = clust, value = centre, -formant) %>%
                    spread(key = formant, value = centre) %>%
                    mutate_all(as.numeric)
                
                list(
                    "classifications" = classifications,
                    "centres" = centres
                )
            
            }
        )
    )

# Re-map mclust's [arbitrary] returned cluster numbers of cross-comparable groups
three_mappings <- . %>%
    mutate(
        oclust = clust,
        clust  = case_when(
            oclust == 1 ~ 1,
            oclust == 2 ~ 3,
            TRUE ~ 2
        )
    )

mclust_best_results[[2]]$classifications <- mclust_best_results[[2]]$classifications %>% three_mappings
mclust_best_results[[2]]$centres <- mclust_best_results[[2]]$centres %>% three_mappings
mclust_best_results[[6]]$classifications <- mclust_best_results[[6]]$classifications %>% three_mappings
mclust_best_results[[6]]$centres <- mclust_best_results[[6]]$centres %>% three_mappings

# have to do 4 forest and 4 praat differently :(
four_f_mappings <- . %>%
    mutate(
        oclust = clust,
        clust  = case_when(
            oclust == 2 ~ 1,
            oclust == 1 ~ 2,
            oclust == 3 ~ 3,
            TRUE ~ 4
        )
    )

mclust_best_results[[3]]$classifications <- mclust_best_results[[3]]$classifications %>% four_f_mappings
mclust_best_results[[3]]$centres <- mclust_best_results[[3]]$centres %>% four_f_mappings

four_p_mappings <- . %>%
    mutate(
        oclust = clust,
        clust  = case_when(
            oclust == 1 ~ 1,
            oclust == 2 ~ 4,
            oclust == 4 ~ 3,
            TRUE ~ 2
        )
    )

mclust_best_results[[7]]$classifications <- mclust_best_results[[7]]$classifications %>% four_p_mappings
mclust_best_results[[7]]$centres <- mclust_best_results[[7]]$centres %>% four_p_mappings

# have to do 5 forest and 5 praat differently too :(
five_f_mappings <- . %>%
    mutate(
        oclust = clust,
        clust = case_when(
            oclust == 1 ~ 1,
            oclust == 2 ~ 3,
            oclust == 5 ~ 2,
            oclust == 3 ~ 5,
            TRUE ~ 4
        )
    )

mclust_best_results[[4]]$classifications <- mclust_best_results[[4]]$classifications %>% five_f_mappings
mclust_best_results[[4]]$centres <- mclust_best_results[[4]]$centres %>% five_f_mappings

five_p_mappings <- . %>%
    mutate(
        oclust = clust,
        clust = case_when(
            oclust == 4 ~ 1,
            oclust == 3 ~ 2,
            oclust == 5 ~ 3,
            oclust == 1 ~ 4,
            TRUE ~ 5
        )
    )

mclust_best_results[[8]]$classifications <- mclust_best_results[[8]]$classifications %>% five_p_mappings
mclust_best_results[[8]]$centres <- mclust_best_results[[8]]$centres %>% five_p_mappings

```

```{r Setup comparison with annotators, message=FALSE, include=FALSE}

vowel_levels <- c(
    "a", "æ", "ɐ", "ɑ", "ɒ",
    "ə", "ɛ", "ɘ", "ɚ", "ɜ",
    "e", "ε", "i", "ĭ", "ɪ", "ɨ",
    "o", "ɔ", "u", "ʉ", "ʊ", "ʌ"
)

tally_pipeline <- . %>%
    group_by(G, tracker, clust, base_vowel) %>%
    summarise(n_vowels = n()) %>%
    ungroup %>%
    tidyr::complete(G, tracker, clust, base_vowel, fill = list(n_vowels = 0)) %>%
    group_by(G, tracker, clust) %>% 
    mutate(
        pc_vowels   = n_vowels / sum(n_vowels) * 100,
        clust_total = sum(n_vowels)
    ) %>% 
    ungroup %>%
    mutate(
        # clust      = paste(clust, " (n = ", clust_total, ")", sep = ""),
        base_vowel = factor(base_vowel, levels = vowel_levels)
    )

plot_vowel_tallies <- function(df) {
    df %>%
    mutate(clust = paste(max(df$G), "G", clust, sep = "")) %>% 
    ggplot(aes(x = base_vowel, y = pc_vowels)) +
    geom_bar(stat = "identity", aes(fill = tracker), position = "dodge", color = "black") +
    facet_wrap(~ clust, scales = "free_x", ncol = 1) +
    scale_y_continuous(limits = c(0, 100)) +
    ylab("Percent label usage within cluster") +
    xlab("IPA label used by annotators") + 
    scale_fill_manual(values = c("white", "grey"))
    # + ggtitle(paste0("G = ", max(df$G)))
}

mclust_best_tallies <- lapply(mclust_best_results, function(fit) { tally_pipeline(fit$classifications) } )

```

### 2-cluster and 3-cluster models

```{r Plot 2- and 3-cluster models, warning=FALSE, fig.width=12, fig.height=6}

grid.arrange(
    plot_mclust_formants(mclust_best_results[[5]], group_colors, group_shapes),
    mclust_best_tallies[[5]] %>% plot_vowel_tallies(),
    ncol = 2,
    layout_matrix = rbind(
        c(1,1,1,NA,NA,NA),
        c(1,1,1,NA,NA,NA),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,NA,NA,NA),
        c(1,1,1,NA,NA,NA)
    )
)


grid.arrange(
    plot_mclust_formants(mclust_best_results[[6]], group_colors, group_shapes),
    mclust_best_tallies[[6]] %>% plot_vowel_tallies(),
    ncol = 2,
    layout_matrix = rbind(
        c(1,1,1,NA,NA,NA),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,NA,NA,NA)
    )
)

```

### 4-cluster and 5-cluster models

```{r Plot 4- and 5-cluster models, warning=FALSE, fig.width=12, fig.height=7}

grid.arrange(
    plot_mclust_formants(mclust_best_results[[7]], group_colors, group_shapes),
    mclust_best_tallies[[7]] %>% plot_vowel_tallies(),
    ncol = 2,
    layout_matrix = rbind(
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2)
    )
)

grid.arrange(
    plot_mclust_formants(mclust_best_results[[8]], group_colors, group_shapes),
    mclust_best_tallies[[8]] %>% plot_vowel_tallies(),
    ncol = 2,
    layout_matrix = rbind(
        # c(1,1,1,NA,NA,NA),
        # c(1,1,1,NA,NA,NA),
        c(NA,NA,NA,2,2,2),
        c(NA,NA,NA,2,2,2),
        c(NA,NA,NA,2,2,2),
        c(NA,NA,NA,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(1,1,1,2,2,2),
        c(NA,NA,NA,2,2,2),
        c(NA,NA,NA,2,2,2),
        c(NA,NA,NA,2,2,2),
        c(NA,NA,NA,2,2,2)
        # c(1,1,1,NA,NA,NA),
        # c(1,1,1,NA,NA,NA)
    )
)

```

# Comparison across optimal models and human annotators

```{r Most frequently-assigned IPA labels}
mclust_best_tallies[5:8] %>%
    bind_rows() %>%
    filter(pc_vowels > 10) %>%
    select(-tracker) %>%
    mutate(pc_vowels = round(pc_vowels, digits = 0))
```

## Distribution of non-tonic schwas

```{r Non-tonic schwas, message=FALSE}
mclust_best_results[5:8] %>%
    map_df(~ .$classifications) %>%
    left_join(
        tibble(
            G = c(2,3,4,5),
            clust = c(2,2,4,5),
            base_vowel = "ə"
        ),
        .
    ) %>%
    mutate(is_tonic = vowel_num == 1) %>%
    group_by(G, clust, is_tonic) %>%
    tally() %>%
    group_by(G, clust) %>%
    mutate(pc = n / sum(n) * 100)

```


# Distribution of contexts within vowel categories

```{r Compute distribution of contexts, include=FALSE}
consonants <- read_csv(here("data/external/consonants.csv"))
vowels     <- read_csv(here("data/external/vowels.csv"))
headwords  <- read_csv(here("data/external/headwords.csv"))

precluster_annotations <- 
    vowels_med_analysis %>%
    group_by(source_file, rep_num, vowel_num) %>%
    select(annotator) %>% 
    nest() %>%
    mutate(token_annotators = map_chr(data, ~ pull(., "annotator") %>% paste0(collapse = ", "))) %>%
    unnest() %>%
    ungroup() %>%
    # To solve. What to do about tokens with not exactly 2 annotators (some 1, some more than 2); very rare, though
    filter(str_count(token_annotators, ",") == 1)

const_groups <- expand.grid(
    c1 = consonants$const,
    c2 = consonants$const,
    stringsAsFactors = FALSE) %>%
    mutate(const_ctx = paste(c1, c2, sep = "_")) %>%
    left_join(rename(consonants, c1 = const, c1_group = const_place))%>%
    left_join(rename(consonants, c2 = const, c2_group = const_place)) %>%
    mutate(const_group = paste(c1_group, c2_group, sep = "_")) %>%
    select(const_ctx, const_group)

const_dists_analysis <- 
    mclust_best_results[[7]]$classifications %>%
    left_join(precluster_annotations) %>%
    filter(!is.na(token_annotators)) %>% 
    # See problem above. Separate will fail if not exactly 2 pieces to separate into.
    separate(token_annotators, into = c("a1", "a2"), sep = ", ") %>% 
    mutate(
        other_annot = ifelse(annotator != a1, a1, a2) 
    ) %>%
    left_join(vowels_med_analysis %>% select(source_file, rep_num, vowel_num, annotator, const_ctx)) %>%
    left_join(vowels_med_analysis %>% select(source_file, rep_num, vowel_num, other_annot = annotator, other_vowel = base_vowel, other_ctx = const_ctx)) %>%
    left_join(
        tibble(
            clust = c(1, 2, 3, 4),
            gmm_group = c("central", "low", "front", "back")
        )
    ) %>%
    left_join(rename(vowels, a1_vgroup = vowel_group)) %>%
    left_join(rename(vowels, other_vowel = base_vowel, a2_vgroup = vowel_group)) %>% 
    filter(gmm_group == a1_vgroup | a1_vgroup == a2_vgroup) %>%
    left_join(const_groups) %>%
    left_join(rename(const_groups, other_ctx = const_ctx, a2_const_group = const_group)) %>%
    filter(const_group == a2_const_group) %>%
    left_join(mutate(headwords, headword = str_to_lower(headword))) %>% 
    separate(const_group, into = c("const_left", "const_right"), sep = "_") %>%
    mutate_at(vars("gmm_group", "const_left", "const_right"), funs(str_to_title))

const_dists_plot <-
    const_dists_analysis %>% 
    group_by(gmm_group, const_left, const_right) %>%
    summarise(
        n_tokens = n(),
        n_types  = unique(headword) %>% length()
    ) %>%
    group_by(gmm_group) %>%
    mutate(
        n_vowels = sum(n_tokens),
        pc_token = round(n_tokens / n_vowels * 100, 1),
        text     = paste(pc_token, "%\n (", n_types, ")", sep = ""),
        vgroup_label = paste(gmm_group, " (n = ", n_vowels, ")", sep = "")
    )
```

```{r Plot context distributions, fig.width=8, fig.height=8}
const_dists_plot %>% 
    ggplot(aes(x = const_right, y = const_left)) +
    geom_tile(aes(fill = pc_token), color = "black") +
    geom_text(aes(label = text)) +
    facet_wrap(~ vgroup_label, scales = "free") +
    scale_fill_gradient(low = "#ffffff", high = "#969696") +
    xlab("Following consonant") +
    ylab("Preceding consonant")
```

# Distribution of vowel categories across vowel positions

```{r Plot position distributions, fig.width=8}
const_dists_analysis %>% 
    group_by(vowel_num, gmm_group) %>%
    summarise(
        n_tokens = n(),
        n_types  = unique(headword) %>% length()
    ) %>%
    filter(vowel_num <= 4) %>% 
    group_by(vowel_num) %>% 
    mutate(
        n_vowels = sum(n_tokens),
        pc_token = round(n_tokens / n_vowels * 100, 1),
        text     = paste(pc_token, "%\n (", n_tokens, ")", sep = ""),
        vgroup_label = paste(gmm_group, " (n = ", n_vowels, ")", sep = "")
    ) %>%
    ungroup %>% 
    mutate(
        bar_label = paste("(", n_types, ")", sep = ""),
        vowel_num = paste("M", vowel_num, " (n = ", n_vowels, ")", sep = "")
    ) %>%
    ggplot(aes(x = as.factor(gmm_group), y = pc_token)) +
    geom_bar(stat = "identity", fill = "white", color = "black", width = 0.8) +
    geom_text(aes(label = bar_label), vjust = -0.6, y = 0)+
    facet_wrap(~ vowel_num, ncol = 5, scales = "free_x") +
    xlab("Medial vowel position in word") +
    ylab("Proportion of occurrence within position") +
    scale_y_continuous(limits = c(0, 65)) +
    theme(
        panel.grid.major = element_line(colour = "#f0f0f0"),
        axis.text.x = element_text(angle = 45, hjust = 1)
    )

```


<style type="text/css">
h1, h2, h3, h4, h5 { font-size:1.1em }
</style>
